Assessing spatial correlations in pulsar timing arrays with the noise-marginalized optimal statistic

### 2017-08-04 manuscript/manuscript.tex
added more references

\begin{abstract}
Supermassive black hole binaries (SMBHBs) are predicted to form a high-redshift cosmological population of low-frequency gravitational wave sources which added together produce a stochastic gravitational wave background (GWB) potentially measurable via pulsar timing. Comparing the actual and predicted times-of-arrival (TOAs) of radio pulses beamed by millisecond pulsars allows us to search for and place limits on the stochastic background. While most estimating of GWB significance is done via Bayesian analysis, here we use a frequentist estimator called the noise-marginalized optimal statistic to assess the significance of a GWB present in simulated PTA data. We find \ki{complete}

@@

Low-frequency gravitational waves (GWs) with frequencies of 
$10^{-9} - 10^{-7} \; \mathrm{Hz}$ can be observed with pulsar timing arrays (PTAs) 
composed of millisecond pulsars (MSPs) \sv{citation}. 
The dominant astrophysical source in this frequency range is an isotropic stochastic 
gravitational wave background (GWB) 
made up of the incoherent superposition of GWs from inspiraling 
supermassive black hole binaries (SMBHBs) 
\citep{1995ApJ...446..543R, 2003ApJ...583..616J, 2003ApJ...590..691W}. 
By monitoring the ultra-regular periodic emission from these pulsars using radio telescopes, 
we can probe the dynamics of the spacetime through which the pulses travel. 
This is done by searching for correlations in the pulsar timing residuals, 
which measure the difference between the expected and observed pulse times of arrival.

The primary data analysis technique used by PTAs is Bayesian inference 
to compare the probabilities of various models for the pulsar timing residuals, 
including one where the residuals contain evidence of an isotropic GW stochastic background \sv{add citation}. 
Bayesian inference is a powerful technique because it allows us to compare arbitrarily complicated models. 
However, running a full Bayesian analysis is computationally intensive, 
particularly when searching for evidence of dipole spatial correlations -- 
``the smoking gun'' of a stochastic GWB.

The significance of a stochastic GWB can also be assessed using the 
optimal statistic, a frequentist estimator for the GWB amplitude \sv{add citation}. 
It is an important complement to Bayesian detection techniques 
because it provides an independent detection procedure, 
and it is significantly faster to compute. 
In particular, different spatial correlations can be tested with the optimal statistic 
in less than a second, 
whereas a Bayesian analysis with spatial correlations requires 
several weeks to run on a supercomputing cluster.

@@

As introduced by \citet{abc+2009}, 
the optimal statistic is derived by analytically maximizing 
the PTA likelihood function in the weak-signal regime. 
We begin by finding the timing residuals for a given pulsar ${\bf{r}}_I$, which can be written as 
\begin{equation}
	{\bf{r}}_I = M {\bf{\epsilon}}_I + F {\bf{a}}_I + U {\bf{j}}_I + {\bf{n}}_I \,,\\
\end{equation}
where $M$ is the design matrix, 
$\epsilon_I$ is a vector of the timing model parameter uncertainties, 
${\bf{a}}_I$ are the red noise parameters describing both individual pulsars' red noise 
and common red noise in all pulsars, 
${\bf{j}}_I$ are white noise parameters describing frequency-correlated white noise, 
and ${\bf{n}}_I$ are white noise parameters describing uncorrelated white noise. 
The likelihood function is constructed from the residuals as given by
\begin{equation}
	p({\bf{r}}|\phi) = \frac{\exp[-\frac{1}{2} {\bf{r}}^T \Sigma^{-1}(\phi) {\bf{r}} ]}{\sqrt{\mathrm{det}[2\pi \Sigma(\phi)]}} \,,
\end{equation}
where $\bf{r}$ is an array of the timing residuals 
from the $M$ pulsars in the PTA,
\begin{equation}
	\bf{r} = \left[ \begin{array}{c} {\bf{r}}_1 \\ {\bf{r}}_2 \\ \vdots \\ {\bf{r}}_M \end{array} \right] \,,
\end{equation}
$\phi$ is the set of pulsar noise parameters, 
and $\Sigma$ is the covariance matrix of the residuals. 
We can write $\Sigma$ in terms of 
autocovariance and cross-covariance matrices ${\bf{P}}_I$ and ${\bf{S}}_{IJ}$, respectively, as
\begin{equation}
	\Sigma = \left[ \begin{array}{cccc} {\bf{P}}_1 & {\bf{S}}_{12} & \hdots & {\bf{S}}_{1M}  \\
							{\bf{S}}_{21} & {\bf{P}}_2 & \hdots & {\bf{S}}_{2M} \\
							\vdots & \vdots & \ddots & \vdots \\
							{\bf{S}}_{M1} & {\bf{S}}_{M2} & \hdots & {\bf{P}}_M \end{array} \right] \,,
\end{equation}
where
\begin{eqnarray}
	{\bf{P}}_I &=& \left\langle {\bf{r}}_I {\bf{r}}_I^T \right\rangle \,, \\
	{\bf{S}}_{IJ} &=& \left. \left\langle {\bf{r}}_I {\bf{r}}_J^T \right\rangle \right|_{I \neq J} \,.
\end{eqnarray}
In the frequency domain, these can be written as
\begin{eqnarray}
	\left\langle {\bf{r}}_I {\bf{r}}_I^T \right\rangle_{ij} &=& {\bf{R}}_I \left[ \int_{-\infty}^\infty df \; e^{2\pi i f \tau_{ij}} P_I(f) \right] {\bf{R}}_I^T \,, \\
	\left. \left\langle {\bf{r}}_I {\bf{r}}_J^T \right\rangle \right|_{ij} &=& \chi_{IJ} {\bf{R}}_I \left[ \int_{-\infty}^\infty df \; e^{2\pi i f \tau_{ij}} P_\mathrm{gw}(f) \right] {\bf{R}}_J^T \,,
\end{eqnarray}
where $i,j$ index the matrix elements, $\tau_{ij} \equiv |t_i - t_j|$ is the time interval between 
two observations, 
$\mathbf{R}$ is the timing model fitting matrix, 
$\chi_{IJ}$ is the overlap reduction function (ORF) that describes the spatial correlation between two pulsars' residuals, 
$P_\mathrm{gw}(f)$ is the GW power spectrum, 
and $P_I(f)$ is the power spectrum for the $I$th pulsar.

@@

We used these three techniques to compute the optimal statistic for simulated data sets. 
We constructed simulated data sets using the same procedure as in \citet{tlb+2017}. 
The simulated PTAs consisted of 10 MSPs with 
total observation times and noise properties chosen to be 
similar to the properties of the EPTA (see Table~\ref{tab:sim})  \sv{add citation}. 
We also injected a stochastic GWB amplitude with $\Agw = 5\times10^{-15}$.
\begin{table}[htb]
	\setlength{\tabcolsep}{5pt}
	\caption{Pulsar parameters used in simulated PTA}
	\begin{center}
	\begin{tabular}{ccccc}
		\hline\hline
    		Pulsar	& $T_\mathrm{obs}$ (yrs) & $\sigma_w$ ($\mu\mathrm{s}$) & $\log_{10}A_\mathrm{red}$ & $\gamma_\mathrm{red}$ \\
		\hline
		J0613$-$0200 & 16.054 & 1.58 & -13.90 & 3.18 \\
		J0751+1807 & 17.606 & 2.60 & -14.14 & 2.58 \\
		J1012+5307 & 16.831 & 1.47 & -13.09 & 1.65 \\
		J1640+2224 & 16.735 & 1.99 & -13.24 & 0.03 \\
		J1643$-$1224 & 17.300 & 1.65 & -18.56 & 4.04 \\
		J1713+0747 & 17.657 & 0.26 & -14.90 & 4.85 \\
		J1744$-$1134 & 17.250 & 0.65 & -13.60 & 2.00 \\
		J1857+0943 & 17.310 & 1.51 & -16.00 & 1.35 \\
		J1909$-$3744 & 9.379 & 0.12 & -13.99 & 2.06 \\
		J2145$-$0750 & 17.161 & 1.19 & -13.87 & 4.02 \\
    		\hline\hline
	\end{tabular}
	\end{center}
	\label{tab:sim}
\end{table}

>>>>

\begin{abstract}
Supermassive black hole binaries (SMBHBs) are predicted to form a 
high-redshift cosmological population of low-frequency gravitational wave sources 
which added together produce a stochastic gravitational wave background (GWB) 
that can be observed by pulsar timing arrays (PTAs). 
Comparing the actual and predicted times-of-arrival (TOAs) of radio pulses beamed by 
millisecond pulsars allows us to search for and place limits on the stochastic background. 
While most estimating of GWB significance is done via Bayesian analysis, 
here we use a frequentist estimator called the noise-marginalized optimal statistic 
to assess the significance of a GWB present in simulated PTA data. We find \ki{complete}

@@

Low-frequency gravitational waves (GWs) with frequencies of 
$10^{-9} - 10^{-7} \; \mathrm{Hz}$ can be observed with pulsar timing arrays (PTAs) 
composed of millisecond pulsars (MSPs) \cite{hd1983,fb1990}. 
The dominant astrophysical source in this frequency range is an isotropic stochastic 
gravitational wave background (GWB) 
made up of the incoherent superposition of GWs from inspiraling 
supermassive black hole binaries (SMBHBs) 
\citep{1995ApJ...446..543R, 2003ApJ...583..616J, 2003ApJ...590..691W}. 
By monitoring the ultra-regular periodic emission from these pulsars using radio telescopes, 
we can probe the dynamics of the spacetime through which the pulses travel. 
This is done by searching for correlations in the pulsar timing residuals, 
which measure the difference between the expected and observed pulse times of arrival. 
Current upper-limits on the stochastic background from PTAs are approaching 
theoretical predictions for the stochastic GWB \citep{PPTA2013,EPTA2015,NANOGrav9yr}

The primary data analysis technique used by PTAs is Bayesian inference 
to compare the probabilities of various models for the pulsar timing residuals, 
including one where the residuals contain evidence of an isotropic GW stochastic background 
\citep{vlm+2009,lah+2013}. 
Bayesian inference is a powerful technique because it allows us to compare arbitrarily complicated models. 
However, running a full Bayesian analysis is computationally intensive, 
particularly when searching for evidence of dipole spatial correlations -- 
``the smoking gun'' of a stochastic GWB.

The significance of a stochastic GWB can also be assessed using the 
optimal statistic, a frequentist estimator for the GWB amplitude \citep{abc+2009,demorest+2013,ccs+2015}. 
It is an important complement to Bayesian detection techniques 
because it provides an independent detection procedure, 
and it is significantly faster to compute. 
In particular, different spatial correlations can be tested with the optimal statistic 
in less than a second, 
whereas a Bayesian analysis with spatial correlations requires 
several weeks to run on a supercomputing cluster.

@@

\sv{Justin, please bring notation up to date}
As introduced by \citet{abc+2009}, 
the optimal statistic is derived by analytically maximizing 
the PTA likelihood function in the weak-signal regime. 
We begin by finding the timing residuals for a given pulsar ${\bf{r}}_I$, which can be written as 
\begin{equation}
	{\bf{r}}_I = M {\bf{\epsilon}}_I + F {\bf{a}}_I + U {\bf{j}}_I + {\bf{n}}_I \,,\\
\end{equation}
where $M$ is the design matrix, 
$\epsilon_I$ is a vector of the timing model parameter uncertainties, 
${\bf{a}}_I$ are the red noise parameters describing both individual pulsars' red noise 
and common red noise in all pulsars, 
${\bf{j}}_I$ are white noise parameters describing frequency-correlated white noise, 
and ${\bf{n}}_I$ are white noise parameters describing uncorrelated white noise. 
The likelihood function is constructed from the residuals as given by
\begin{equation}
	p({\bf{r}}|\phi) = \frac{\exp[-\frac{1}{2} {\bf{r}}^T \Sigma^{-1}(\phi) {\bf{r}} ]}{\sqrt{\mathrm{det}[2\pi \Sigma(\phi)]}} \,,
\end{equation}
where $\bf{r}$ is an array of the timing residuals 
from the $M$ pulsars in the PTA,
\begin{equation}
	\bf{r} = \left[ \begin{array}{c} {\bf{r}}_1 \\ {\bf{r}}_2 \\ \vdots \\ {\bf{r}}_M \end{array} \right] \,,
\end{equation}
$\phi$ is the set of pulsar noise parameters, 
and $\Sigma$ is the covariance matrix of the residuals. 
We can write $\Sigma$ in terms of 
autocovariance and cross-covariance matrices ${\bf{P}}_I$ and ${\bf{S}}_{IJ}$, respectively, as
\begin{equation}
	\Sigma = \left[ \begin{array}{cccc} {\bf{P}}_1 & {\bf{S}}_{12} & \hdots & {\bf{S}}_{1M}  \\
							{\bf{S}}_{21} & {\bf{P}}_2 & \hdots & {\bf{S}}_{2M} \\
							\vdots & \vdots & \ddots & \vdots \\
							{\bf{S}}_{M1} & {\bf{S}}_{M2} & \hdots & {\bf{P}}_M \end{array} \right] \,,
\end{equation}
where
\begin{eqnarray}
	{\bf{P}}_I &=& \left\langle {\bf{r}}_I {\bf{r}}_I^T \right\rangle \,, \\
	{\bf{S}}_{IJ} &=& \left. \left\langle {\bf{r}}_I {\bf{r}}_J^T \right\rangle \right|_{I \neq J} \,.
\end{eqnarray}
In the frequency domain, these can be written as
\begin{eqnarray}
	\left\langle {\bf{r}}_I {\bf{r}}_I^T \right\rangle_{ij} &=& {\bf{R}}_I \left[ \int_{-\infty}^\infty df \; e^{2\pi i f \tau_{ij}} P_I(f) \right] {\bf{R}}_I^T \,, \\
	\left. \left\langle {\bf{r}}_I {\bf{r}}_J^T \right\rangle \right|_{ij} &=& \chi_{IJ} {\bf{R}}_I \left[ \int_{-\infty}^\infty df \; e^{2\pi i f \tau_{ij}} P_\mathrm{gw}(f) \right] {\bf{R}}_J^T \,,
\end{eqnarray}
where $i,j$ index the matrix elements, $\tau_{ij} \equiv |t_i - t_j|$ is the time interval between 
two observations, 
$\mathbf{R}$ is the timing model fitting matrix, 
$\chi_{IJ}$ is the overlap reduction function (ORF) that describes the spatial correlation between two pulsars' residuals, 
$P_\mathrm{gw}(f)$ is the GW power spectrum, 
and $P_I(f)$ is the power spectrum for the $I$th pulsar.

@@

We used these three techniques to compute the optimal statistic for simulated data sets. 
We constructed simulated data sets using the same procedure as in \citet{tlb+2017}. 
The simulated PTAs consisted of 10 MSPs with 
total observation times and noise properties chosen to be 
similar to the properties of the EPTA \cite{caballero+2016} (see Table~\ref{tab:sim}). 
We also injected a stochastic GWB amplitude with $\Agw = 5\times10^{-15}$.
\begin{table}[htb]
	\setlength{\tabcolsep}{5pt}
	\caption{Pulsar parameters used in simulated PTA}
	\begin{center}
	\begin{tabular}{ccccc}
		\hline\hline
    		Pulsar	& $T_\mathrm{obs}$ (yrs) & $\sigma_w$ ($\mu\mathrm{s}$) & $\log_{10}A_\mathrm{red}$ & $\gamma_\mathrm{red}$ \\
		\hline
		J0613$-$0200 & 16.054 & 1.58 & -13.90 & 3.18 \\
		J0751+1807 & 17.606 & 2.60 & -14.14 & 2.58 \\
		J1012+5307 & 16.831 & 1.47 & -13.09 & 1.65 \\
		J1640+2224 & 16.735 & 1.99 & -13.24 & 0.03 \\
		J1643$-$1224 & 17.300 & 1.65 & -18.56 & 4.04 \\
		J1713+0747 & 17.657 & 0.26 & -14.90 & 4.85 \\
		J1744$-$1134 & 17.250 & 0.65 & -13.60 & 2.00 \\
		J1857+0943 & 17.310 & 1.51 & -16.00 & 1.35 \\
		J1909$-$3744 & 9.379 & 0.12 & -13.99 & 2.06 \\
		J2145$-$0750 & 17.161 & 1.19 & -13.87 & 4.02 \\
    		\hline\hline
	\end{tabular}
	\end{center}
	\label{tab:sim}
\end{table}
